---
title: "NABCA IBRC"
author: "Tina Cheng"
date: "7/25/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)
```

```{r}
library(tidyverse)
library(flextable)

theme_set(theme_bw() +
            theme(panel.grid = element_blank(), 
                  text = element_text(size = 14)))
```



```{r}
USresults <- read_csv(here::here("data/US_CA_survey718871_25July2022.csv"))
MXresults <- read_csv(here::here("data/MX_results_2022July25.csv"))
threatNum <- read_csv(here::here("data/ThreatNum.csv")) 
```

## ABOUT

NABCA State of the Bats data for USA, CAN, and MEX are composed of reviewer estimates for population size & trend and the scope and severity of threats. 



```{r}
USresults_char <- USresults %>% mutate_all(as.character)
MXresults_char <- MXresults %>% mutate_all(as.character)
token_lookup <- USresults_char %>% bind_rows(MXresults_char) %>% distinct(token, firstname, lastname) 
#   filter(!is.na(lastname))

#USA Data

#data munging
USlong <- USresults %>% 
  mutate_all(as.character) %>%
  pivot_longer(cols = `popSize.sz_min.`:`otherSev.user10_conf.`,
               names_to = "category") %>% 
  dplyr::select(id, token, sppCode, category, value, cntry) %>% 
  mutate(cntry = case_when(
    cntry == "US" ~ "USA",
    cntry == "CA" ~ "CAN"
  ))

USparse <- USlong %>% 
  separate(category, into = c("category", "statistic"), sep = "_") %>% 
  separate(category, into = c("category", "subcategory"), sep = "\\.") %>% 
  mutate(metric = case_when(
    grepl("Scope", category) ~ "scope",
    grepl("Sev", category) ~ "severity",
    grepl("Size", category) ~ "size",
    grepl("Trend", category) ~ "trend"
  ),
  category = gsub("Scope", "", category),
  category = gsub("Sev", "", category),
  category = gsub("Size", "", category),
  category = gsub("Trend", "", category)) %>% 
  mutate(statistic = gsub("\\.", "", statistic)) 

#dataframe of mean values only for each category


#MEX data

#data munging
MXlong <- MXresults %>% 
  filter(id != 533) %>% #weird trend (7500)
  mutate_all(as.character) %>%
  pivot_longer(cols = `popSize[sz_min]`:`otherSev[user10_conf]`,
               names_to = "category") %>% 
  dplyr::select(id, token, sppCode, category, value, cntry) %>% 
  mutate(cntry = "MEX")

#dataframe of mean values only for each category
MXparse <- MXlong %>% 
  separate(category, into = c("category", "statistic"), sep = "_") %>% 
  separate(category, into = c("category", "subcategory"), sep = "\\[") %>% 
  mutate(metric = case_when(
    grepl("Scope", category) ~ "scope",
    grepl("Sev", category) ~ "severity",
    grepl("Size", category) ~ "size",
    grepl("Trend", category) ~ "trend"
  ),
  category = gsub("Scope", "", category),
  category = gsub("Sev", "", category),
  category = gsub("Size", "", category),
  category = gsub("Trend", "", category)) %>% 
  mutate(statistic = gsub("\\]", "", statistic)) %>% 
  mutate(subcategory = gsub("\\]", "", subcategory)) 

long <- USparse %>% 
  bind_rows(MXparse) %>% 
  filter(!is.na(sppCode)) #%>% 
  # filter(token %in% token_lookup$token) #filter down to only tokens that have a name associated

means <- USparse %>% 
  bind_rows(MXparse) %>% 
  filter(!is.na(sppCode)) #%>% 
  # filter(statistic == "mean") %>% 
  # filter(token %in% token_lookup$token)

wider <- USparse %>% 
  bind_rows(MXparse) %>% 
  pivot_wider(id_cols = c(token, sppCode, category, metric, subcategory, cntry), names_from = statistic, values_from = value) %>% 
  mutate(interval_width = as.numeric(`max`) - as.numeric(`min`),
         `conf` = as.numeric(`conf`)) %>% 
  rename(conf = `conf`,
         min = `min`,
         max = `max`) #%>% 
  # filter(token %in% token_lookup$token)
```

```{r}
#Pop 

popSummary <- means %>% 
  filter(!is.na(sppCode)) %>% 
  filter(category == "pop") %>% 
  group_by(sppCode, category, subcategory, metric, cntry) %>% 
  mutate(value = as.numeric(value)) %>% 
  dplyr::summarise(meanValue = mean(value, na.rm = T), 
                   n = n(),
                   sd = sd(value, na.rm = T)) %>% 
  mutate(trend_cat = case_when(
    subcategory == "tr" & n < 3 ~ "insufficient reviewers", 
    subcategory == "tr" & n >= 3 & meanValue < 0 ~ "decreasing",
    subcategory == "tr" & n >= 3 & meanValue == 0 ~ "stable",
    subcategory == "tr" & n >= 3 & meanValue > 0 ~ "increasing"
  ),
  trend_cat_decreasing  = case_when(
    trend_cat == "decreasing" & meanValue < 0 & meanValue >= -10 ~ "slight (1-10%)", 
    trend_cat == "decreasing" & meanValue < 10 & meanValue >= -30 ~ "moderate (11-30%)", 
    trend_cat == "decreasing" & meanValue < 30 & meanValue >= -70 ~ "serious (31-70%)", 
    trend_cat == "decreasing" & meanValue < 70 & meanValue >= -100 ~ "extreme (71-100%)" 
  )) 
```

```{r}
trend <- popSummary %>% 
  filter(subcategory == "tr") %>% 
  group_by(cntry, trend_cat) %>% 
  summarise(n = n()) %>%
  ungroup() %>% 
  add_row(cntry = "MEX", trend_cat = "not reviewed", n = 51) %>% #n: took the number of spp reported by IUCN (140) and subtracted the number reviewed (89)
  group_by(cntry) %>% 
  mutate(cntry_total = sum(n),
         prop = n/cntry_total)
```

```{r}

# Threats

assignImpact <-  function(dat, scope, severity){
  
  scope <- enquo(scope)
  severity <- enquo(severity)
  
  dat %>% 
    mutate_at(vars(!!scope, !!severity), round) %>% 
    mutate(impactScore = case_when(!!scope <= 10 & !!severity >= 1 ~ "low",
                                   !!scope >= 11 & !!scope <= 30 & !!severity >= 1 & !!severity <= 30 ~ "low",
                                   !!scope >= 11 & !!scope <= 30 & !!severity >= 31 ~ "medium",
                                   !!scope >= 31 & !!severity <= 10 ~ "low",
                                   !!scope >= 31 & !!scope <= 70 & !!severity >= 11 & !!severity <= 30 ~ "medium",
                                   !!scope >= 31 & !!scope <= 70 & !!severity >= 31 ~ "high",
                                   !!scope >= 71 & !!severity >= 11 & !!severity <= 30 ~ "medium",
                                   !!scope >= 71 & !!severity >= 31 & !!severity <= 70 ~ "high",
                                   !!scope >= 71 & !!severity >= 71 ~ "very high",
                                   !!scope == 0 | is.na(!!scope) & !!severity == 0 | is.na(!!severity) ~ "negligible"))
}

threats <- means %>% 
  filter(!is.na(metric)) %>% 
  filter(category != "pop") %>% 
  mutate(value = as.numeric(value)) %>% 
  pivot_wider(id_cols = c("sppCode", "category", "subcategory", "token", "cntry"), names_from = "metric", values_from = "value") %>% 
  mutate(scope = as.numeric(scope),
         severity = as.numeric(severity)) %>%
  mutate(impact_numeric = ((scope/100) * (severity/100))) %>% 
  left_join(threatNum, by = c("category" = "Q_group", "subcategory" = "Q_sub")) 

#Threats summary
threatSum_step1 <- threats %>% 
  filter(!is.na(sppCode) & !is.na(Threat)) %>% 
  mutate_if(is.numeric, replace_na, replace = 0) %>% 
  group_by(sppCode, token, cntry) %>% 
  dplyr::summarise(sumImpact = sum(impact_numeric, na.rm = T)) %>% 
  left_join(token_lookup, by = c("token"="token")) %>% 
  unite("name", c(firstname, lastname), sep= " ", remove = T)
  
  
threatSum <- threatSum_step1 %>% filter(sumImpact<=1) %>% 
  ungroup() %>% 
  group_by(sppCode, cntry) %>% 
  summarise(meanImpact = mean(sumImpact),
            sd = sd(sumImpact),
            n=n()) %>% 
  filter(n>1)

outlier_fig <- ggplot() +
  geom_point(data = threatSum_step1, aes(y = sppCode, x = sumImpact, color = name, text = name), , alpha = 0.7, size = 3) + 
  xlab("Sum of Threat Impacts") +
  ylab("")+
  ggtitle("Sum of Threat Impacts for Each Reviewer for Each Species") +
  theme(axis.text.y = element_text(hjust = 1, size = 6), legend.text=element_text(size=10), plot.title = element_text(size = 12))

outlier_figly <- ggplotly(outlier_fig, width = 1000, height = 1000, 
                  tooltip = c("text", "y"))
outlier_figly

saveWidget(myfig, "/Users/ngoodby/Downloads/threat_ass.html")
# %>% 
#   assignImpact(., meanScope, meanSeverity) %>%
#   mutate(impactScore = factor(impactScore, levels = c("negligible", "low", "medium", "high", "very high"))) %>% 
#   left_join(threatNum, by = c("category" = "Q_group", "subcategory" = "Q_sub")) %>% 
#   filter(!is.na(Threat))

```

```{r}
library("scales")

pop_numbers <- popSummary %>% filter(n>1) %>% 
  pivot_wider(names_from = metric, values_from = meanValue) %>% ungroup() %>% select(sppCode, cntry, size, trend) %>%
  group_by(sppCode, cntry) %>% 
  summarise_all(list(~ sum(., na.rm = T))) %>% 
  mutate(previous_pop = size - size*0.01*trend)



# ggplot(pop_numbers, aes(size, trend, color = cntry)) + 
#   geom_point(alpha = 0.5, size = 4) +
#   geom_text(aes(label=sppCode)) +
#   scale_x_continuous(labels = comma_format(big.mark = ",",
#                                            decimal.mark = "."))
#                      limits = c(0, 10e6))
```

```{r, fig.height = 4, fig.width = 4}
overall_pop <- popSummary %>% filter(metric == "trend") %>% group_by(sppCode) %>% summarise(overall_trend = mean(meanValue)) %>% 
  arrange(desc(overall_trend)) 

spp_order <- overall_pop$sppCode
wider$sppCode <- factor(wider$sppCode, levels = spp_order)

wider %>% filter(subcategory == "tr" & !is.na(sppCode)) %>% group_by(sppCode, cntry) %>% summarise(min = mean(as.numeric(min)),
                                                 max = mean(as.numeric(max)),
                                                 mean = mean(as.numeric(mean))) %>% 
  ggplot(aes(y = sppCode, x = mean, color = cntry)) +
  geom_vline(xintercept = 0, color = "grey") + 
  geom_point() +
  geom_errorbar(aes(y = sppCode, xmin = min, xmax = max), width = 0) +
  geom_point(data = overall_pop, aes(y = sppCode, x = overall_trend), color = "red", alpha = 0.4, size = 3) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), text = element_text(size = 6))
```

```{r, fig.height = 4, fig.width = 4}
library(plotly)
library(htmlwidgets)
#impact = scope x severity
impact <- threatSum %>% 
  mutate(min = (meanImpact - sd),
        max = (meanImpact + sd))

overall_impact <- impact %>% group_by(sppCode) %>% summarise(overall_mean_impact = mean(meanImpact),
                                                             overall_min_impact = mean(min),
                                                             overall_max_impact = mean(max)) %>% 
  arrange(overall_mean_impact)

spp_order <- overall_impact$sppCode
impact$sppCode <- factor(impact$sppCode, levels = spp_order)
overall_impact$sppCode <- factor(overall_impact$sppCode, levels = spp_order)

fig <- ggplot() +
  geom_vline(xintercept = 1, linetype = "dashed") + 
  geom_errorbar(data = overall_impact, aes(y = sppCode, xmin = overall_min_impact, xmax = overall_max_impact), width = 1, color = "grey")+
  # geom_errorbar(aes(y = sppCode, xmin = min, xmax = max, color = cntry), width = 0) 
  geom_point(data = impact, aes(y = sppCode, x = meanImpact, color = cntry), alpha = 0.7, size = 4) +
  geom_point(data = overall_impact, aes(y = sppCode, x = overall_mean_impact), color = 'purple', size =2, shape = 4) +
  xlab("Threat Impact") +
  ylab("")+
  ggtitle("Overall Threat Impact on North American Bat Species (impact = scope x severity)") +
  theme(axis.text.y = element_text(hjust = 1, size = 6), legend.text=element_text(size=10), plot.title = element_text(size = 12))

myfig <- ggplotly(fig, width = 1000, height = 1000, 
                  tooltip = c("x", "y"))
myfig

saveWidget(myfig, "/Users/ngoodby/Downloads/fig.html")
```

```{r}
threatSum %>% 
  group_by(cntry, sppCode) %>% 
  summarize(totalDec=sum(meanImpact, na.rm = T)) %>% 
  arrange(-totalDec)
```

```{r}
threats %>% 
  select(sppCode, subcategory, impact_numeric) %>% 
  mutate(impact_numeric=case_when(is.na(impact_numeric) ~ 0,
                                  T~impact_numeric)) %>% 
  filter(!is.na(sppCode),
         !is.na(subcategory),
         !is.na(impact_numeric)) %>% 
  kmeans(centers=10)
```

```{r}
threatSum %>% 
  group_by(cntry, sppCode) %>% 
  summarize(totalDec=sum(meanImpact, na.rm = T)) %>% 
  arrange(-totalDec) %>% 
  group_by(cntry) %>% 
  summarize(n_gt1 = sum(totalDec>1),
            n_tn1 = sum(totalDec<1))
```

```{r}
threatSum %>% 
  group_by(cntry, sppCode) %>% 
  summarize(totalDec=sum(meanImpact, na.rm = T)) %>% 
  arrange(-totalDec) %>% 
  group_by(cntry) %>%
  ggplot(aes(y=totalDec, fill= cntry, group=cntry))+
  geom_histogram()
```

```{r fig.height=4.5, fig.width=4}
threatSum %>%
  group_by(cntry, sppCode, Threat) %>% 
  summarize(totalCat_Dec=sum(meanImpact, na.rm = T)) %>% 
  group_by(cntry, sppCode) %>% 
  mutate(rank=rank(totalCat_Dec)) %>% 
  group_by(cntry, Threat) %>% 
  summarize(n=n(),
            totalRank=sum(rank),
            avgRank=totalRank/n) %>% 
  arrange(cntry,totalRank) %>% 
  ggplot(aes(x=Threat, y=avgRank, group=cntry, fill=cntry))+
  geom_col()+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```



## Population Stats

**Narrative:**

- Mexico: species are largely declining but declines are slight (less than 10%)
- USA: a majority of species are stable or decreasing; of those decreasing, 13% are extreme declines
- CAN: a majority of species (65%) lacked enough reviewers to make an assessment; of those species that were reviewed, all are decreasing; of those that are decreasing, 17% are extreme and 33% are serious declines

These trends show that experts are worried and further highlights why we need empirical evidence to determine what is really happening with these populations.

**Notes:**

- Mean pop trend values do not reflect uncertainty both as indicated by the reviewer (lower and upper estimates, confidence rating)
- There are a number of pop trend estimates whose upper and lower values overlap zero, indicating that the trend is not significantly different from zero (Supplemental Figures 1-3)

### Methods

I took the mean value of population trend among reviewers per species per country. I categorized the mean trend for each species (per country) as:

- not reviewed: this was only the case for MEX; I subtracted the number of spp reviewed (89) by the total number of spp in MEX (140; IUCN Red List)
- insufficient reviewers: less than 3 reviewers per species per country
- increasing (>0)
- stable (=0)
- decreasing (<0)


I further categorized decreasing species by the severity of decline (NatureServe) as:

- slight (1-10%)
- moderate (11-30%)
- serious (31-70%)
- extreme (71-100%)

For each country, I show the proportion of species in each trend category (Figure 1; Table 1) and by severity of decline (Figure 2; Table 2).


Table 1. Proportion of species by population trend.

```{r}
trend %>% 
  mutate(trend_display = paste0(round(prop, 2)*100, "% (", n, "/", cntry_total, ")"),
         trend_cat = factor(trend_cat, levels = c("increasing", "stable", "decreasing", "insufficient reviewers", "not reviewed"))) %>% 
  arrange(trend_cat) %>% 
  pivot_wider(id_cols = trend_cat, names_from = cntry, values_from = trend_display) %>% 
  rename(`Population Trend` = trend_cat) %>% 
  flextable() %>% 
  autofit()
```

Table 2. Proportion of species by severity of decline

```{r}
popSummary %>% 
  filter(trend_cat == "decreasing") %>% 
  mutate(trend_cat_decreasing = factor(trend_cat_decreasing, levels = c("extreme (71-100%)", "serious (31-70%)", "moderate (11-30%)", "slight (1-10%)")),
         cntry = factor(cntry, levels = c("CAN", "USA", "MEX"))) %>% 
  group_by(cntry, trend_cat_decreasing) %>% 
  summarise(n = n()) %>% 
  ungroup() %>% 
  group_by(cntry) %>% 
  mutate(cntry_total = sum(n),
         prop = n/cntry_total) %>% 
  mutate(trend_display = paste0(round(prop, 2)*100, "% (", n, "/", cntry_total, ")")) %>% 
  arrange(trend_cat_decreasing) %>% 
  pivot_wider(id_cols = trend_cat_decreasing, names_from = cntry, values_from = trend_display) %>% 
  rename(`Severity of Decline` = trend_cat_decreasing) %>% 
  flextable() %>% 
  autofit()
```

```{r, eval = F, fig.height = 5, fig.width = 10}
USsppRank_popSize <- popSummary %>% 
  filter(cntry %in% c("USA", "CAN")) %>% 
filter(category == "pop" & subcategory == "sz") %>% 
  arrange(meanValue) %>% 
  distinct(sppCode) %>% pull(sppCode)

p1 <- USmeans %>% 
  filter(!is.na(sppCode)) %>% 
  filter(category == "pop" & subcategory == "sz") %>% 
  mutate(sppCode = factor(sppCode, levels = USsppRank_popSize)) %>% 
  ggplot(aes(x = sppCode, y = as.numeric(value))) +
  geom_point() +
  geom_point(data = USpopSummary %>% 
               filter(category == "pop" & subcategory == "sz") %>% 
               mutate(sppCode = factor(sppCode, levels = USsppRank_popSize)),
             aes(x = sppCode, y = meanValue), color = "red", size = 2) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  xlab("") +
  ylab("Population Size") +
  scale_y_log10() +
  ggtitle("USA")

MXsppRank_popSize <- MXpopSummary %>% 
  filter(category == "pop" & subcategory == "sz") %>% 
  arrange(meanValue) %>% 
  distinct(sppCode) %>% pull(sppCode)


p2 <- MXmeans %>% 
  filter(!is.na(sppCode)) %>% 
  filter(category == "pop" & subcategory == "sz") %>% 
  mutate(sppCode = factor(sppCode, levels = MXsppRank_popSize)) %>% 
  ggplot(aes(x = sppCode, y = as.numeric(value))) +
  geom_point() +
  geom_point(data = MXpopSummary %>% 
               filter(category == "pop" & subcategory == "sz") %>% 
               mutate(sppCode = factor(sppCode, levels = MXsppRank_popSize)),
             aes(x = sppCode, y = meanValue), color = "red", size = 2) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  xlab("") +
  ylab("Population Size") +
  scale_y_log10() +
  ggtitle("MEX")

cowplot::plot_grid(p1, p2)
```



```{r, fig.cap = "Figure 1. Proportion of species by population trend. Species are categorized as increasing (trend > 0), stable (trend = 0), or decreasing (trend < 0) based on the mean value of trends solicited from expert reviewers. For species with fewer than three reviewers, no trend was calculated and that species is indicated as having `insufficient reviewers.` For Mexico, there were 51 (out of 140) species that were not reviewed."}

trend %>% 
  mutate(trend_cat = factor(trend_cat, levels = c("increasing", "stable", "decreasing", "insufficient reviewers", "not reviewed")),
         cntry = factor(cntry, levels = c("CAN", "USA", "MEX"))) %>% 
  ggplot() +
  geom_bar(aes(x = cntry, y = prop, fill = trend_cat), stat = "identity", position = "stack") +
  scale_fill_manual(values = c("#2c7bb6", "#abd9e9", "#d7191c", "#838383", "#cfcfcf"), "Population \n Trend") +
  coord_flip() +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(nrow = 3)) +
  xlab("") +
  ylab("Proportion of Species")

```

make country abbrevs three letters

```{r, fig.cap = "Figure 2. Proportion of species by serverity of decline. Species are categorized by decline severity (slight, moderate, serious, or extreme) based on NatureServe criteria."}

popSummary %>% 
  filter(trend_cat == "decreasing") %>% 
  mutate(trend_cat_decreasing = factor(trend_cat_decreasing, levels = c("extreme (71-100%)", "serious (31-70%)", "moderate (11-30%)", "slight (1-10%)")),
         cntry = factor(cntry, levels = c("CAN", "USA", "MEX"))) %>% 
  ggplot() +
  geom_bar(aes(x = cntry, fill = trend_cat_decreasing), position = "fill") +
  scale_fill_manual(values = c("#B30000", "#FF3333", "#FF6666", "#FFB3B3"), "Severity \nof Decline") +
  coord_flip() +
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(nrow = 2)) +
  xlab("") +
  ylab("Proportion of Species")
```



Difficult to establish trends for Mexico just based on mean values of trend b/c many of them are really close to zero, so if we were to calculate the 95% CI, they would likely overlap zero.



```{r, fig.height = 10, fig.width = 8}

plot_trend <- function(cntry_sel){
  # sppRank_popTrend <- popSummary %>% 
  #   filter(cntry == cntry_sel) %>% 
  #   filter(category == "pop" & subcategory == "tr") %>% 
  #   arrange(meanValue) %>% 
  #   distinct(sppCode) %>% pull(sppCode)
  
  wider %>% 
    filter(subcategory == "tr" & cntry == cntry_sel) %>% 
    filter(mean < 500) %>% 
    ggplot(aes(x = sppCode, y = as.numeric(mean))) +
    geom_hline(yintercept = 0, color = "grey") +
    geom_point() +
    geom_errorbar(aes(x = sppCode, ymin = as.numeric(min), ymax = as.numeric(max)), width = 0) +
    geom_point(data = popSummary %>% filter(cntry == cntry_sel & metric == "trend"), aes(x = sppCode, y = meanValue), color = "red", alpha = 0.4, size = 3) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    ggtitle(cntry_sel)
  
  
  # means %>% 
  #   filter(cntry == cntry_sel) %>% 
  #   filter(category == "pop" & subcategory == "tr") %>% 
  #   mutate(sppCode = factor(sppCode, levels = sppRank_popTrend)) %>% 
  #   ggplot(aes(x = sppCode, y = as.numeric(value))) +
  #   geom_hline(yintercept = 0, color = "grey") +
  #   geom_point(size = 2, alpha = 0.6) +
  #   geom_point(data = popSummary %>% filter(cntry == cntry_sel),
  #              aes(x = sppCode, y = meanValue), color = "red", size = 2) +
  #   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  #   xlab("") +
  #   ylab("Population Trend (%)") +
  #   ylim(-100, 100) +
  #   ggtitle(cntry_sel)
  
}
```


```{r, fig.height = 4, fig.width = 10, fig.width = 8, fig.height = 4, fig.cap = "Supplemental Figure 1. Population trend by species for Mexico. Population trends submitted by each reviewer (black points = mean, black lines = upper and lower estimates) are shown with mean values (red points) overlaid."}
plot_trend("MEX")
```

```{r, fig.height = 4, fig.width = 10, , fig.cap = "Supplemental Figure 1. Population trend by species for USA. Population trends submitted by each reviewer (black points = mean, black lines = upper and lower estimates) are shown with mean values (red points) overlaid."}
plot_trend("USA")
```

```{r, fig.height = 4, fig.width = 8, , fig.cap = "Supplemental Figure 1. Population trend by species for Canada. Population trends submitted by each reviewer (black points = mean, black lines = upper and lower estimates) are shown with mean values (red points) overlaid."}
plot_trend("CAN")
```

## Threats

*What are the threats of highest concern among experts?*

- We know WNS and wind are stressors, but what do experts think about chronic stressors?

- How do we manage species in a multi-threat context?


### Methods

For each threat level II, I calculated the mean estimate for scope and severity across reviewers for each species. I then categorized "impact" as:

- negligible: scope and severity is zero or NA
- low
- medium
- high
- very high

according to this chart for scope and severity:

!(/Users/tcheng/Box/- Missions & Programs/Research & Development/NABat/NABCA/NABCA SoB Report/Analysis_IBRC/www/natureserve_impact.png)




```{r, eval = F, fig.height = 6, fig.width = 12}
threats <- unique(threatNum$Q_group)



p1 <- USsummary %>% 
  filter(!is.na(Threat)) %>% 
  filter()
ggplot(aes(x = threat_abbrev, y = meanValue, color = sppCode)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 14)) +
  theme(legend.position = "none") +
  ylim(0, 100) +
  ggtitle("USA")

p2 <- MXsummary %>% 
  filter(!is.na(Threat)) %>% 
  ggplot(aes(x = threat_abbrev, y = meanValue, color = sppCode)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 14)) +
  theme(legend.position = "none") +
  ylim(0, 100) +
  ggtitle("MEX")


cowplot::plot_grid(p1, p2)
```


*Which threat impacts the most species?*


```{r}
#sum up threats by spp (to get rid of sub-categories)
threatAll <- threatSum %>% 
  filter(!is.na(Threat)) %>% 
  group_by(sppCode, cntry, Threat, threat_abbrev) %>% 
  summarise(totImpact = sum(meanImpact, na.rm = T),
            meanImpact = mean(meanImpact, na.rm = T),
            meanScopeAll = mean(meanScope, na.rm = T),
            meanSeverityAll = mean(meanSeverity, na.rm = T),
            totScopeAll = sum(meanScope, na.rm = T),
            totSeverityAll = sum(meanSeverity, na.rm = T)) %>% 
  assignImpact(., meanScopeAll, meanSeverityAll) %>% 
  rename(meanImpactScore = impactScore) %>% 
  assignImpact(., totScopeAll, totSeverityAll) %>% 
  rename(totImpactScore = impactScore) %>% 
  mutate(totImpactScore = factor(totImpactScore, levels = c("very high", "high", "medium", "low", "negligible")),
         meanImpactScore = factor(meanImpactScore, levels = c("very high", "high", "medium", "low", "negligible"))
  ) 

#number of species experiencing a threat by impact level
threatSppCount <- threatAll %>% 
  filter(!is.na(Threat)) %>% 
  group_by(Threat, threat_abbrev, cntry, totImpactScore) %>% 
  summarise(noSpp = n())

meanThreatSppCount <- threatAll %>% 
  filter(!is.na(Threat)) %>% 
  group_by(Threat, threat_abbrev, cntry, meanImpactScore) %>% 
  summarise(noSpp = n())

# threatAll %>% 
#   ggplot(aes(x = Threat, fill = totImpactScore)) +
#   geom_bar(position = "fill") +
#   facet_wrap(~cntry) +
#   theme(axis.text.x = element_text(angle = 90)) +
#   coord_flip()
```

*MEAN THREAT: Take the mean of sub-categories within a threat group*

```{r, fig.height = 10, fig.width = 12}
#count the number of species per threat per impact score

plotMeanThreat <- function(cntry_sel){
  
  threatSub <- meanThreatSppCount %>% 
    filter(cntry == cntry_sel) 
  
  # threatRank <- threatSub %>% complete(threat_abbrev, meanImpactScore) %>%
  #   mutate(noSpp = ifelse(is.na(noSpp), 0, noSpp),
  #          impScoreNum = as.numeric(meanImpactScore)) %>% filter(impScoreNum == max(impScoreNum)) %>%  arrange(meanImpactScore, noSpp) %>% pull(threat_abbrev) 
  
  threatSub %>% 
    # mutate(threat_abbrev = factor(threat_abbrev, levels = threatRank)) %>% 
    filter(meanImpactScore %in% c("very high", "high", "medium", "low")) %>% 
    ggplot() +
    geom_bar(aes(x = threat_abbrev, y = noSpp), position = "dodge", stat = "identity") +
    facet_wrap(~meanImpactScore, nrow = 1) +
    coord_flip() +
    theme(axis.text.x = element_text(angle = 90)) +
    ggtitle(cntry_sel) +
    ylab("Number of Species") +
    xlab("Threat")
}

p1 <- plotMeanThreat("MEX")

p2 <- plotMeanThreat("USA")

p3 <- plotMeanThreat("CAN")

cowplot::plot_grid(p1, p2, p3, ncol = 1)
```

*TOTAL THREAT: Take the sum of sub-categories within a threat group*

```{r, fig.height = 10, fig.width = 12}
#count the number of species per threat per impact score

plotThreat <- function(cntry_sel){
  
  threatSub <- threatSppCount %>% 
    filter(cntry == cntry_sel) 
  
  threatRank <- threatSub %>% complete(threat_abbrev, totImpactScore) %>% filter(totImpactScore == "very high") %>% mutate(noSpp = ifelse(is.na(noSpp), 0, noSpp)) %>%  arrange(noSpp) %>% pull(threat_abbrev) 
  
  threatSub %>% 
    mutate(threat_abbrev = factor(threat_abbrev, levels = threatRank)) %>% 
    filter(totImpactScore %in% c("very high", "high", "medium", "low")) %>% 
    ggplot() +
    geom_bar(aes(x = threat_abbrev, y = noSpp), position = "dodge", stat = "identity") +
    facet_wrap(~totImpactScore, nrow = 1) +
    coord_flip() +
    theme(axis.text.x = element_text(angle = 90)) +
    ggtitle(cntry_sel) +
    ylab("Number of Species") +
    xlab("Threat Level I")
}

p1 <- plotThreat("MEX")

p2 <- plotThreat("USA")

p3 <- plotThreat("CAN")

cowplot::plot_grid(p1, p2, p3, ncol = 1)
```

*Climate Change: Sub-categories*

Version 1: Number of species by impact level (very high, high, medium, low) in each climate change sub-category 

```{r, fig.height = 10, fig.width = 12}
#deeper look at climate change
plotSubThreat <- function(cntry_sel, threat_sel){
  
  threatSub <- threatSum %>% 
    filter(Threat == threat_sel) %>% 
    filter(cntry == cntry_sel) %>% 
    group_by(Threat, subT, impactScore) %>% 
    summarise(noSpp = n())
  
  subThreatRank <- threatSub %>% filter(Threat == threat_sel) %>%  complete(subT, impactScore) %>% filter(impactScore == "very high") %>% mutate(noSpp = ifelse(is.na(noSpp), 0, noSpp)) %>%  arrange(noSpp) %>% pull(subT) 
  
  threatSub %>% 
    filter(Threat == threat_sel) %>% 
    mutate(subT = factor(subT, levels = subThreatRank),
           impactScore = factor(impactScore, levels = c("very high", "high", "medium", "low", "negligible"))) %>% 
    #filter(impactScore %in% c("very high", "high", "medium", "low")) %>% 
    ggplot() +
    geom_bar(aes(x = subT, y = noSpp), position = "dodge", stat = "identity") +
    facet_wrap(~impactScore, nrow = 1) +
    coord_flip() +
    theme(axis.text.x = element_text(angle = 90)) +
    ggtitle(cntry_sel) +
    ylab("Number of Species") +
    xlab("Threat Level II")
}

p1 <- plotSubThreat("MEX", "Climate Change")
p2 <- plotSubThreat("USA", "Climate Change")
p3 <- plotSubThreat("CAN", "Climate Change")

cowplot::plot_grid(p1, p2, p3, ncol = 1)

# threatSum %>% 
#   filter(Threat == "Climate Change") %>% 
#   ggplot(aes(x = subT, y = meanImpact, group = sppCode, fill = impactScore)) +
#   geom_bar(stat = "identity", position = "dodge") +
#   theme(
#         axis.text.x = element_text(angle = 90)) +
#   facet_wrap(~cntry, scales = "free", ncol = 1) +
#   coord_flip()

```
Version 2: Proportion of species by impact level (very high, high, medium, low) in each climate change sub-category 

```{r, fig.height = 10, fig.width = 12}
plotSubThreat_v2 <- function(cntry_sel, threat_sel){
  threatSub <- threatSum %>% 
    filter(Threat == threat_sel) %>% 
    filter(cntry == cntry_sel) 
  
  threatSub %>% 
    filter(Threat == threat_sel) %>% 
    filter(impactScore %in% c("very high", "high", "medium", "low", "negligible")) %>% 
    ggplot() +
    geom_bar(aes(x = subT, fill = impactScore), position = "fill") +
    scale_fill_manual(values = c("#B30000", "#FF3333", "#FF6666", "#FFB3B3"), "Severity \nof Decline") +
    coord_flip() +
    theme(axis.text.x = element_text(angle = 90), legend.position = "bottom") +
    ggtitle(cntry_sel) +
    ylab("Proportion of Species") +
    xlab("Threat Level II")
}

p1 <- plotSubThreat_v2("MEX", "Climate Change") + theme(legend.position = "none")
p2 <- plotSubThreat_v2("USA", "Climate Change") + theme(legend.position = "none")
p3 <- plotSubThreat_v2("CAN", "Climate Change") + theme(legend.position = "none")

p_leg <- cowplot::get_legend(p1)

p <- cowplot::plot_grid(p1, p2, p3, ncol = 1)

cowplot::plot_grid(p, p_leg, ncol = 1, rel_heights = c(1, 0.1))
```

**Energy Production & Mining**

Version 1: Number of species by impact level (very high, high, medium, low) in each energy sub-category 

```{r, fig.height = 10, fig.width = 12}



p1 <- plotSubThreat("MEX", "Energy Production & Mining")
p2 <- plotSubThreat("USA", "Energy Production & Mining")
p3 <- plotSubThreat("CAN", "Energy Production & Mining")

cowplot::plot_grid(p1, p2, p3, ncol = 1)



```

Version 2: Proportion of species by impact level (very high, high, medium, low) in each energy sub-category 

```{r, fig.height = 10, fig.width = 12}

p1 <- plotSubThreat_v2("MEX", "Energy Production & Mining") + theme(legend.position = "none")
p2 <- plotSubThreat_v2("USA", "Energy Production & Mining") + theme(legend.position = "none")
p3 <- plotSubThreat_v2("CAN", "Energy Production & Mining") + theme(legend.position = "none")

p_leg <- cowplot::get_legend(p1)

p <- cowplot::plot_grid(p1, p2, p3, ncol = 1)

cowplot::plot_grid(p, p_leg, ncol = 1, rel_heights = c(1, 0.1))

```


```{r, eval = F}
threatSum %>% 
  filter(Threat == "Climate Change") %>% 
  filter(cntry == "MEX") %>% 
  #gather(c(scope:impact_numeric), key = "type", value = "value") %>% 
  
  ggplot() +
  geom_hline(yintercept = 55) +
  geom_bar(aes(x = subT, y = meanImpact, fill = subT, group = sppCode), position = "dodge", stat = "identity") +
  coord_flip() +
  theme(axis.text.x = element_text(angle = 90),
        legend.position = "bottom") +
  ylim(0, 100) +
  ylab("Number of Species") +
  theme(legend.position = "none") +
  xlab("Threat")

threats %>% 
  
  #pivot_longer(cols = c("scope", "severity", "impact_numeric"), names_to = "type", values_to = "value") %>% 
  #group_by(Threat, subT, cntry, type, sppCode) %>% 
  #summarise(meanValue = mean(value, na.rm = T),
  #  n = n())# %>% 
  filter(Threat == "Climate Change") %>% 
  filter(cntry == "MEX") %>% 
  ggplot() +
  geom_point(aes(x = scope, y = severity, color = subT)) +
  facet_wrap(~subT) +
  theme(legend.position = "none")

```


*Which species are most impacted?*

```{r, fig.height = 20, fig.width = 6}


plot_species_threat <- function(cntry_sel){
  sppRank_threats <- threatAll %>% filter(cntry == cntry_sel) %>% group_by(sppCode) %>% summarise(overallImpact = sum(totImpact, na.rm = T)) %>% arrange(overallImpact) %>% pull(sppCode)
  
  threatAll %>% 
    filter(cntry == cntry_sel) %>% 
    mutate(sppCode = factor(sppCode, levels = sppRank_threats)) %>% 
    ggplot(aes(x = sppCode, y = totImpact, fill = Threat)) +
    geom_bar(position = "stack", stat = "identity") +
    coord_flip() +
    facet_wrap(~cntry) +
    theme(legend.position = "bottom") +
    ggtitle(cntry_sel)
}

# leg <- cowplot::get_legend(p1 + guides(color = guide_legend(nrow = 1)) +
#     theme(legend.position = "bottom"))
# 
# p1 <- plot_species_threat("MEX") + theme(legend.position = "none") 
# p2 <- plot_species_threat("USA") + theme(legend.position = "none") 
# p3 <- plot_species_threat("CAN") + theme(legend.position = "none") 
# p <- cowplot::plot_grid(p1, p2, p3, nrow = 3)
# 
# cowplot::plot_grid(p, leg, ncol = 1, rel_heights = c(2, 0.4))
```

```{r, fig.height = 16, fig.width = 6}

plot_species_threat("MEX")
```

```{r, fig.height = 8, fig.width = 6}

plot_species_threat("USA")
```

```{r, fig.height = 6, fig.width = 6}

plot_species_threat("CAN")
```

```{r, eval = FALSE, fig.height = 10, fig.width = 20}



threats %>% 
  filter(cntry == "USA") %>% 
  filter(sppCode == "MYTHY") %>% 
  filter(!is.na(scope)) %>% 
  group_by(category) %>% 
  summarise(n = n(),
            no_reviewers = length(unique(token)))

# tmyth <- threats %>% 
#   filter(cntry == "USA") %>% 
#   filter(sppCode == "MYTHY") %>% 
#   #dplyr::select(-impact_numeric) %>% 
#   filter(category == "cli") %>% 
#   left_join(threatNum, by = c("category" = "Q_group", "subcategory" = "Q_sub"))
#   
# 
# tmyth %>% 
#   gather(c(scope:impact_numeric), key = "type", value = "value") %>% 
#   ggplot(aes(x = subT, y = value, group = token, fill = type)) +
#   geom_bar(stat = "identity", position = "dodge") +
#   #facet_wrap(~Threat, scales = "free") +
#   facet_wrap(~type) +
#   coord_flip() +
#   ylim(0, 100) +
#   theme(legend.position = "bottom")

threatSum %>% 
  filter(cntry == "USA") %>% 
  filter(sppCode == "MYTHY") %>% 
  filter(category == "cli") %>% 
  ggplot(aes(x = subT, y = meanImpact)) +
  geom_bar(stat = "identity") +
  geom_point(data = tmyth, aes(x = subT, y = impact_numeric), color = "red", size = 4) +
  coord_flip()

```

```{r}
knitr::knit_exit()
```


### PCA

```{r, eval = F}


USpca <- USsummary %>% 
  filter(!is.na(Threat)) %>% 
  mutate(meanValue = as.numeric(meanValue)) %>% 
  pivot_wider(id_cols = c("sppCode"), names_from = c(category, subcategory), values_from = meanValue) %>% 
  ungroup() %>% 
  dplyr::select(-sppCode) %>% 
  mutate_all(funs(replace_na(.,0))) %>% 
  as.data.frame()
row.names(USpca) <- unique(USsummary$sppCode)


USpcaResults <- prcomp(USpca)
USpcaResults$rotation <- -1*USpcaResults$rotation
#USpcaResults$rotation
USpcaResults$x <- -1*USpcaResults$x
head(USpcaResults$x)
biplot(USpcaResults)

```



